Estimating SARS-CoV-2 transmission in England from randomly sampled prevalence surveys

### 2023-11-29 ATACCC.tex
ATACCC todos

First, I will explain how biomarkers are typically modelled, both for SARS-CoV-2 and historically in the biostatistical literature (\cref{ATACCC:sec:background}).
Next, I describe the model of \textcite{hakkiOnset} in detail, which applies these methods to the ATACCC data (\cref{ATACCC:sec:hakki}) and before I modifying its observation model (\cref{ATACCC:sec:observation-modification})\todo{think if some justification is needed here, and if so how to do it - remember Hakki is a published and peer-reviewed analysis}.
This modification is based on an earlier analysis using a preliminary version of the dataset, described in \cref{ATACCC:sec:original-analysis}.
Finally, I convert the model and parameter estimates of the biomarker's trajectory to a duration of PCR positivity (\cref{ATACCC:sec:duration}).
I will refine this estimate in future chapters before its use in estimating transmission.

@@

The noisy observations are modelled by assuming each measurement of log viral load is a realisation of an iid Normally-distributed random variable error with mean 0 and standard deviation $\sigma_v$.
\todo{Paul/Dani: is this phrasing "a realisation of [a] random variable" OK?}
The measurements are left-censored at the limit of detection, $\vlod$, meaning that any observation which would give a log viral load of less than $\vlod$ is instead observed as negative.
False positives and negatives are modelled by allowing a test failure with probability $\rho$.
Test failures are modelled as producing a log viral load reading drawn from $\N(x_0, \sigma_0^2)$ where $x_0$ and $\sigma_0^2$ are parameters to be estimated.
These readings are censored in the same way as true readings.
A censored test failure represents a false negative, otherwise the test failure represents a false positive.
\textcite{hakkiOnset} uses a prior for $x_0$ centred on $\vlod$, which encodes a prior belief that false negatives and false positives occur with equal probability.

@@

The original \textcite{hakkiOnset} model has convergence issues (Rhat > 1.07 and ESS < 100) for the parameters of the test failure distribution, $\sigma_0$ and $\rho$.
The simplified model I proposed in \cref{ATACCC:sec:observation-modification} solved this issue, meaning that this analysis converged.
\todo[inline]{Add traceplots}

@@

For individual-level parameters, both models had convergence issues in places (Rhat > 1.4 and ESS < 20).
However, this was greatly improved with the simplified model, occurring only for parameters relating to a single individual, number 55.
Individual 55 has an extremely noisy set of observations.
This means that two viral load trajectories fit the data well (shown in \cref{ATACCC:fig:individual-55}).
The first, found by two chains, increases very quickly, peaks 4.0 days after the first test, then declines slowly.
The second, found by the other six chains, is more symmetrical, peaking 7.2 days after the first test.
The population-level parameter estimates agree between the two sets of chains, therefore the issue should not affect any estimaates of interest.
\begin{figure}
  \centering \includegraphics{ATACCC/fit_individual_55}
  \caption[Goodness-of-fit for individual 55]{Fit to individual 55's data coloured by MCMC chain. Two chains favour the quickly increasing viral load trajectory, peaking at day 4.0, the other six favour the slower trajectory, peaking at day 7.2. \label{ATACCC:fig:individual-55}}
\end{figure}

@@

>>>>

First, I will explain how biomarkers are typically modelled, both for SARS-CoV-2 and historically in the biostatistical literature (\cref{ATACCC:sec:background}).
Next, I describe the model of \textcite{hakkiOnset} in detail, which applies these methods to the ATACCC data (\cref{ATACCC:sec:hakki}) and before modifying its observation model to improve the inference (\cref{ATACCC:sec:observation-modification}).
This modification is based on an earlier analysis using a preliminary version of the dataset, described in \cref{ATACCC:sec:original-analysis}.
Finally, I convert the model and parameter estimates of the biomarker's trajectory to a duration of PCR positivity (\cref{ATACCC:sec:duration}).
I will refine this estimate in future chapters before its use in estimating transmission.

@@

The noisy observations are modelled by assuming each measurement of log viral load has an iid error distributed $\N(0, \sigma_v)$.
The measurements are left-censored at the limit of detection, $\vlod$, meaning that any observation which would give a log viral load of less than $\vlod$ is instead observed as negative.
False positives and negatives are modelled by allowing a test failure with probability $\rho$.
Test failures are modelled as producing a log viral load reading drawn from $\N(x_0, \sigma_0^2)$ where $x_0$ and $\sigma_0^2$ are parameters to be estimated.
These readings are censored in the same way as true readings.
A censored test failure represents a false negative, otherwise the test failure represents a false positive.
\textcite{hakkiOnset} uses a prior for $x_0$ centred on $\vlod$, which encodes a prior belief that false negatives and false positives occur with equal probability.

@@

The original \textcite{hakkiOnset} model has convergence issues (Rhat > 1.07 and ESS < 100) for the parameters of the test failure distribution, $\sigma_0$ and $\rho$.
The simplified model I proposed in \cref{ATACCC:sec:observation-modification} solved this issue, meaning that this analysis converged.

@@

For individual-level parameters, both models had convergence issues in places (Rhat > 1.4 and ESS < 20).
However, this was greatly improved with the simplified model, occurring only for parameters relating to a single individual, number 55.
Individual 55 has an extremely noisy set of observations.
This means that two viral load trajectories fit the data well (shown in \cref{ATACCC:fig:individual-55}).
The first, found by two chains, increases very quickly, peaks 4.0 days after the first test, then declines slowly.
The second, found by the other six chains, is more symmetrical, peaking 7.2 days after the first test.
The population-level parameter estimates agree between the two sets of chains, therefore the issue should not affect any estimates of interest.
\begin{figure}
  \centering \includegraphics{ATACCC/fit_individual_55}
  \caption[Goodness-of-fit for individual 55]{Fit to individual 55's data coloured by MCMC chain. Two chains favour the quickly increasing viral load trajectory, peaking at day 4.0, the other six favour the slower trajectory, peaking at day 7.2. \label{ATACCC:fig:individual-55}}
\end{figure}

@@

The negative rate, $\rho$, is estimated to be quite low, 5.3\% (95\% CrI: 3.7--7.3\%).
